---
title: "vegetation_analysis"
author: "Liyenne"
date: "2023-09-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```

```{r include = FALSE} 
#setwd("~/Project Green Arctic/2_malla/1_data/field_data/1_malla_data_analysis") 

require(tidyverse)
require(tibble)

require(ggplot2)
require(ggordiplots)

require(vegan)
require(lme4)
require(glmmTMB)
```

```{r}
SE <- function(x) sd(x)/sqrt(length(x))
```

# Data import
**vegetation data**
```{r echo = FALSE}
vegetation_data <- readr::read_csv('../../2_input/1_raw_data/vegetation_data_raw.csv')

vegetation_data %>% head(10)
```
**dung count data**
```{r}
dung_data <- read.csv("../../3_output/1_data_products/dung_count_data.csv")

dung_data %>% 
  head(10)
```

**meta data**
```{r echo = FALSE}
meta_data <- readr::read_csv('../../2_input/1_raw_data/plot_data_extended.csv')
meta_data %>% head(10)

dnp <- readxl::read_excel("../../2_input/1_raw_data/chem/DNP.xlsx") %>% 
  dplyr::select(code, NO3, NH4, PO4_corrected)

soil_chem <- readr::read_csv("../../2_input/1_raw_data/chem/chem_malla_soil_leaves.csv") %>% 
  filter(type == "soil") %>% 
  dplyr::select(Sample_ID, wN, d15N, wC, d13C, CN_ratio) %>% 
  rename(code = Sample_ID)

meta_data <- meta_data %>% 
  left_join(dnp, by = "code") %>% 
  left_join(soil_chem, by = "code")
```

The code below appends the sum of scats found in the 10 meters closes to each plot to the meta data
```{r}
# add dung counts to vegetation_data
vegetation_data <- dung_data %>% 
  mutate(habitat = case_when(
    distance <= -41 ~ "forest", 
    distance >= -5 & distance <= 5 ~ "treeline", 
    distance >= 45 & distance <= 54 ~ "tundra",
    distance >= 91 ~ "upper_tundra"
  )) %>% 
  na.omit() %>% 
  group_by(transect, habitat) %>% 
  summarise(dung_count = sum(count)) %>% 
  inner_join(vegetation_data, ., by = join_by(transect, habitat)) %>% 
  dplyr::select(code:habitat, dung_count, everything())
  
```

```{r}
vegetation_data_l <- vegetation_data %>% 
  pivot_longer(salix:rock, names_to = "type", values_to = "cover")

vegetation_data_l %>% 
  head(10)

vegetation_data_l$country <- factor(vegetation_data_l$country, levels = c("N", "F"))

vegetation_data_l %>% 
  filter(type == c("salix", "gram")) %>% 
ggplot(aes(habitat, cover)) +
  geom_boxplot(aes(fill = country)) +
    labs(fill = "Grazing regime") +
  scale_fill_manual(values = c("#77BFBC", "#EB956A"), labels = c("Winter", "Year-round")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ type) +
  theme_classic() +
  theme(panel.spacing = unit(.0, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1), 
        strip.background = element_rect(color = "black", size = 1),
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../3_output/2_graphs/veg_covers.svg", height = 3.5, width = 5)
```


# NMDS
```{r}
vegetation_tbl <- vegetation_data %>% # add environmental variables
  full_join(meta_data) %>% 
  select(code:habitat, dung_count, slope, aspect, twi, everything()) %>% 
  column_to_rownames("code") 

vegetation_tbl %>% 
  as.data.frame() %>% 
  head(10)

meta_data <- vegetation_tbl %>% # create meta data 
  mutate(country_habitat = paste0(country, habitat)) %>% 
  select(country, habitat, country_habitat, dung_count, slope, aspect, twi, everything())
```

```{r include = FALSE}
# multivariate (NMDS)
  # scat counts around plots
  # slope
  # forest density (n trees)
  # aspect
  # soil data
  # ndvi?

veg_nmds <- vegetation_data %>% 
  select(code, salix:rock) %>% 
  column_to_rownames("code") %>% 
  metaMDS() %>% 
  scores(tidy = TRUE) %>%
  as_tibble(ronames = "code")
  
species_nmds <- veg_nmds %>% 
  filter(score == "species")

veg_nmds <- veg_nmds %>% 
  filter(score == "sites") %>% 
  select(NMDS1, NMDS2)
  
meta_nmds <- cbind(meta_data, veg_nmds)
``` 

```{r NMDS plot (country), echo = FALSE}
hull_data <- meta_nmds %>% # find convex hulls
  group_by(country) %>% # for each combination of country and habitat
  arrange(NMDS1, NMDS2) %>%
  slice(chull(NMDS1, NMDS2))

ggplot() + 
  geom_point(data = meta_nmds, aes(x = NMDS1, y = NMDS2, shape = habitat, colour = country),size=3) + # add the point markers
  geom_polygon(data=hull_data,aes(x=NMDS1,y=NMDS2, fill = country),alpha=0.30) +
  geom_text(data=species_nmds,aes(x=NMDS1,y=NMDS2,label=label),alpha=0.5) +  # add the species labels
  #geom_text(data = data_scores, aes(x = NMDS1, y = NMDS2, label = plot), vjust=0) +  # add the site labels
  coord_equal() +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(shape = NA)))
```

```{r NMDS plot (country and habitat), echo = FALSE}
hull_data <- meta_nmds %>% # find convex hulls
  group_by(country, habitat) %>% # for each combination of country and habitat
  arrange(NMDS1, NMDS2) %>%
  slice(chull(NMDS1, NMDS2))

ggplot() +
  scale_linetype_manual(values = c("forest" = "solid", "treeline" = "dashed", "tundra" = "dotdash", "upper_tundra" = "dashed")) +
  geom_point(data = meta_nmds, aes(x = NMDS1, y = NMDS2, shape = habitat, colour = country), size = 3) + # add the point markers 
  geom_polygon(data = hull_data, aes(x = NMDS1, y = NMDS2, color = country, fill = country, linetype = habitat), alpha = 0.30) +
  geom_text(data = species_nmds, aes(x = NMDS1, y = NMDS2, label = label), alpha = 0.5) +  # add the species labels
  #geom_text(data = data_scores, aes(x = NMDS1, y = NMDS2, label = plot), vjust=0) +  # add the site labels
  coord_equal() +
  theme_bw() +
  guides(color = guide_legend(override.aes = list(shape = NA)))
```

```{r PEMANOVA}
adonis2(as.matrix(vegetation_data[, 7:18]) ~ country * habitat, vegetation_data, permu = 999, method = "bray")
```

Within group variation found among habitats (forest, treeline, tundra, upper tundra) in Norway and Finland does not differ. Meaning that the outcome of the PERMANOVA is likely reliable. 
```{r}
vegetation_tbl %>% 
  select(salix:rock) %>% 
  as.dist() %>% 
  betadisper(meta_data$country_habitat) %>% 
  permutest()
```

fitting environmental variables
```{r include=FALSE}
nmds <- vegetation_data %>% 
  select(code, salix:rock) %>% 
  column_to_rownames("code") %>% 
  metaMDS()

```


```{r }
ef <- meta_data %>% 
  select(dung_count, slope, aspect, twi, NO3, NH4, PO4_corrected, wN:CN_ratio, country, habitat, transect) %>% 
  envfit(nmds, ., na.rm = TRUE); ef
```

**NMDS with environmental variables**
```{r echo = FALSE}
ef_coord_cont <-  as.data.frame(scores(ef, "vectors")) * ordiArrowMul(ef) # map vectors in plot
ef_coord_cont <- ef_coord_cont[!(row.names(ef_coord_cont) %in% c("slope", "wN", "wC")),]

rownames(ef_coord_cont) <- c("dung", "aspect", "TWI", "nitrate", "ammonium", "phosphate", "dN15", "dC13", "C/N ratio") 

ef_coord_cat = as.data.frame(scores(ef, "factors")) * ordiArrowMul(ef) # map factors in plot
ef_coord_cat <- ef_coord_cat[1:6,]
rownames(ef_coord_cat) <- c("Finland", "Norway", "forest", "treeline", "tundra", "high tundra")

hull_data <- meta_nmds %>% # find convex hulls
  group_by(country) %>% # for each combination of country and habitat
  arrange(NMDS1, NMDS2) %>%
  slice(chull(NMDS1, NMDS2))

ggplot() + 
  geom_point(data = meta_nmds, aes(x = NMDS1, y = NMDS2, shape = habitat, colour = country),size=3) + # add the point markers
  scale_color_manual(values = c("#EB956A", "#77BFBC"), labels = c("Year-round", "Winter")) + 
  geom_polygon(data = hull_data, aes(x = NMDS1, y = NMDS2, fill = country, color = country),alpha=0.30) + 
  scale_fill_manual(values = c("#EB956A", "#77BFBC"), labels = c("Year-round", "Winter")) +
  geom_text(data=species_nmds,aes(x=NMDS1,y=NMDS2,label=label),alpha=0.5) +  # add the species labels
  #geom_text(data = data_scores, aes(x = NMDS1, y = NMDS2, label = plot), vjust=0) +  # add the site labels
  coord_equal() +
  geom_segment(aes(x = 0, y = 0, xend = NMDS1, yend = NMDS2), 
       data = ef_coord_cont, linewidth =1, alpha = 0.5, colour = "grey30") +
     geom_point(data = ef_coord_cat, aes(x = NMDS1, y = NMDS2), 
       shape = "diamond", size = 4, alpha = 0.6, colour = "navy") +
     geom_text(data = ef_coord_cat, aes(x = NMDS1, y = NMDS2+0.08), 
       label = row.names(ef_coord_cat), colour = "navy", fontface = "bold") + 
     geom_text(data = ef_coord_cont, aes(x = NMDS1, y = NMDS2), colour = "grey30", 
       fontface = "bold", label = row.names(ef_coord_cont)) + 
  labs(fill = "Grazing regime", color = "Grazing regime", shape = "Habitat") +
  theme_classic() +
  guides(color = guide_legend(override.aes = list(shape = NA))) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1), 
        strip.background = element_rect(color = "black", size = 1))
```

```{r}
gg_ordisurf(ord = nmds, env.var = meta_data$aspect, binwidth = 5)
```

# type cover
```{r}
vegetation_data[, 7:18] <- (vegetation_data[, 7:18]/100)
vegetation_data$country <- as.factor(vegetation_data$country)
vegetation_data$habitat <- as.factor(vegetation_data$habitat)
vegetation_data$transect <- as.factor(vegetation_data$transect)
```

```{r}
vegetation_data %>% 
  group_by(country) %>% 
  summarise(msalix = mean(salix), salix_se = SE(salix), mgram = mean(gram), gram_se = SE(gram), mforb = mean(forb), forb_se = SE (forb), mbareGr = mean(bareGr), bareGr_se = SE(bareGr))
```

## graminoids
```{r graminoid boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = gram, fill = country)) +
  geom_boxplot() 
```

```{r graminoid LMMM, echo = FALSE}
dist <- fitdist(vegetation_data$gram, "beta")
plot(dist)

fit <- glmmTMB(gram ~ country * habitat + (1 | transect), 
               dispformula = ~ country * habitat,
               family = beta_family(), data = vegetation_data) 
summary(fit)
car::Anova(fit, type = "II")
```

```{r graminoid LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```
```{r graminoid model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```

```{r}
vegetation_data %>% 
  group_by(country) %>% 
  summarise(mean(gram), SE(gram))
```

## forbs
```{r forb boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = forb, fill = country)) +
  geom_boxplot() 
```

```{r forb LMMM, echo = FALSE}
dist <- fitdist(vegetation_data$forb, "exp")
plot(dist)

fit <- glmmTMB(forb ~ country * habitat + (1 | transect), 
               dispformula = ~ country * habitat,
               family = beta_family(), ziformula = ~1, 
               data = vegetation_data) 

summary(fit)
car::Anova(fit, type = "II")
```

```{r forb LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r forb model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```
 
## salix
```{r salix boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = salix, fill = country)) +
  geom_boxplot() 
```

```{r salix LMMM, echo = FALSE}
dist <- fitdist(vegetation_data$salix, "exp")
plot(dist)

fit <- glmmTMB(salix ~ country * habitat + (1 | transect), 
               family = beta_family(), ziformula = ~1, 
               data = vegetation_data) 

summary(fit)
car::Anova(fit, type = "II")
```

```{r salix LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r salix model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```

## B. nana
```{r B. nana boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = Bnana, fill = country)) +
  geom_boxplot() 
```

```{r B. nana LMMM, echo = FALSE}
dist <- fitdist(vegetation_data$Bnana, "exp")
plot(dist)

fit <- vegetation_data %>% 
  filter(habitat != "forest") %>% 
  glmmTMB(Bnana ~ country * habitat + (1 | transect), 
               dispformula = ~ country * habitat,
               family = beta_family(), ziformula = ~1, 
               data = .) 

summary(fit)
car::Anova(fit, type = "II")
```

```{r B. nana LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r B. nana model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput) 
par(mfrow = c(1,2))
plotResiduals(simulationOutput, vegetation_data$country)
plotResiduals(simulationOutput, vegetation_data$habitat) # variation in the forest plots with respect to cover deviates from other groups
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```

## J. communis
```{r J. communis boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = Jcomm, fill = country)) +
  geom_boxplot() 
```

```{r J. communis LMMM, echo = FALSE}
dist <- fitdist(vegetation_data$Jcomm, "exp")
plot(dist)

fit <- glmmTMB(Jcomm ~ country * habitat + (1 | transect), 
               family = beta_family(), ziformula = ~1, 
               data = vegetation_data) 

summary(fit)
car::Anova(fit, type = "II")
```

```{r J. communis LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r J. communis model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```

## deciduous dwarf shrubs
```{r deciduous dwarf shrub boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = dDwarf, fill = country)) +
  geom_boxplot() 
```

```{r deciduous dwarf shrub LMMM, echo = FALSE}
dist <- fitdist(vegetation_data$dDwarf, "exp")
plot(dist)

fit <- glmmTMB(dDwarf ~ country * habitat + (1 | transect), 
               family = beta_family(), ziformula = ~1, 
               data = vegetation_data) 

summary(fit)
car::Anova(fit, type = "II")
```

```{r deciduous dwarf shrub LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r deciduous dwarf shrub model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```

## evergreen dwarf shrubs
```{r evergreen dwarf shrub boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = eDwarf, fill = country)) +
  geom_boxplot() 
```

```{r evergreen dwarf shrub LMMM, echo = FALSE}
dist <- fitdist(vegetation_data$eDwarf, "exp")
plot(dist)

fit <- glmmTMB(eDwarf ~ country * habitat + (1 | transect), 
               family = beta_family(), ziformula = ~1, 
               data = vegetation_data) 

summary(fit)
car::Anova(fit)
```

```{r evergreen dwarf shrub LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r evergreen dwarf shrub model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```

## bryophytes
```{r bryophyte boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = bry, fill = country)) +
  geom_boxplot() 
```

```{r bryophyte LMMM, echo = FALSE}
dist <- fitdist(log(vegetation_data$bry), "norm")
plot(dist)

fit <- glmmTMB(log(bry) ~ country * habitat + (1 | transect), 
               dispformula = ~ country * habitat,
               #family = beta_family(), 
               ziformula = ~1, 
               data = vegetation_data) 

summary(fit)
car::Anova(fit, type = "II")
```

```{r bryophyte LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r bryophyte model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plotResiduals(simulationOutput, vegetation_data$country) #heterogeneity of variance 
plotResiduals(simulationOutput, vegetation_data$habitat)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```

## lichens
```{r lichen boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = lich, fill = country)) +
  geom_boxplot() 
```

```{r lichen LMMM, echo = FALSE}
dist <- fitdist(vegetation_data$lich, "exp")
plot(dist)

fit <- glmmTMB(lich ~ country * habitat, 
               dispformula = ~ country * habitat,
               family = beta_family(), ziformula = ~1, 
               data = vegetation_data) 

summary(fit)
car::Anova(fit)
```

```{r lichen LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r lichen model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```

## bare grounds
```{r bare ground boxplot, echo = FALSE}
vegetation_data %>% 
  na.omit() %>% 
  ggplot(aes(x = habitat, y = bareGr, fill = country)) +
  geom_boxplot() 
```

```{r bare ground LMMM, echo = FALSE}
vegetation_data$bareGr[vegetation_data$bareGr == 0] <- 0.001
  
dist <- fitdist(vegetation_data$bareGr, "exp")
plot(dist)

fit <- glmmTMB(bareGr ~ country, 
               family = beta_family(), ziformula = ~1, 
               data = vegetation_data) 

summary(fit)
car::Anova(fit)
```

```{r bare ground LMMM, echo = FALSE}
emmeans(fit, pairwise ~ country)$contrasts
emmeans(fit, pairwise ~ habitat)$contrasts
emmeans(fit, pairwise ~ country | habitat)$contrasts
```

```{r bare ground model diagnostics}
simulationOutput <- simulateResiduals(fittedModel = fit)
plot(simulationOutput)
testDispersion(simulationOutput); rm(simulationOutput)

fit <- lmer(d15N ~ country + habitat + (1|transect_ID), data = cn_soil)
x <- check_collinearity(fit)
plot(x); rm(x)
```