---
title: "tree_data_analysis"
author: "Liyenne"
date: "2023-09-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(error = TRUE)
```

```{r include = FALSE} 
require(kableExtra)

require(tidyr)
require(dplyr)

require(lme4)
require(lmerTest)
require(lsmeans)
require(MuMIn)
require(AER)
require(MASS)

require(ggplot2)
require(ggpubr)
require(flexplot)
```

# Data
**individual tree data**
```{r include = FALSE}
tree_data <- readr::read_csv('../../3_output/1_data_products/tree_data.csv')
```

```{r echo = FALSE}
tree_data %>% head(10)
```

**transect meta/environmental data**
```{r include = FALSE}
# transect meta data
transect_data <- readr::read_csv('../../3_output/1_data_products/transect_data.csv')

transect_data <- readr::read_csv('../../3_output/1_data_products/dung_data_transect_summary.csv') %>% 
  select(-frequency) %>% 
  rename(habitat = section) %>% 
  full_join(transect_data)
transect_data %>% head(10)
```


```{r include = FALSE}
# tree transect summary data
tree_transect_data <- readr::read_csv('../../3_output/1_data_products/tree_transect_data.csv')
tree_transect_data %>% head(10)
```

**tree data summarized per transect**
```{r include = FALSE}
tree_transect_data_full <- readr::read_csv('../../3_output/1_data_products/tree_transect_data_full.csv')
tree_transect_data_full %>% head(10)

tree_transect_data_full <- tree_transect_data_full %>% 
  left_join(transect_data, by = c("transect" = "transect", "country" = "country", "section" = "habitat"))
```

# Analysis
```{r include = FALSE}
#of trees in all size classes (e.g., seedlings, smaller than 50 cm; saplings, smaller than 2m; adult trees, taller than 2m)
tree_data$country <- factor(tree_data$country, levels = c("Storfjord", "Malla"))

tree_data %>%
  na.omit() %>%
  ggplot(aes(distance, color = country, fill = country)) + 
  geom_histogram(alpha = 0.5, 
                position="identity") +
  geom_vline(aes(xintercept = 0), linetype = 'dashed') +
  facet_wrap(~country)
```


```{r include = FALSE}
#only adult trees (taller than 2m)
tree_data %>%
  filter(class == 'tree') %>%
  ggplot(aes(distance, color = country, fill = country)) + 
  geom_histogram(alpha = 0.5, position="identity") +
  geom_vline(aes(xintercept = 0), linetype = 'dashed') +
  facet_wrap(~country)
```


```{r include = FALSE}
# seedlings and saplings
facet_order <- c("seedling", "sapling", "tree", "dead")
  
tree_data %>% #keep for later
  #filter(class == 'sapling') %>%
  mutate(class = factor(class, levels = facet_order)) %>% 
  ggplot(aes(distance, color = country, fill = country)) + 
  geom_histogram(alpha = 0.5, 
                position="identity") +
  geom_vline(aes(xintercept = 0), linetype = 'dashed') +
  facet_wrap(~class)
```

**numbers of trees**
```{r echo = FALSE}
facet_order <- c("seedling", "sapling", "tree", "dead")

tree_transect_data_full$country <- factor(tree_transect_data_full$country, levels = c("Storfjord", "Malla"))

tree_transect_data_full %>%
  mutate(class = factor(class, levels = facet_order)) %>% 
  #filter(class != "dead" & class != "tree") %>%  
  ggplot(aes(section, n, fill = country)) + 
  geom_boxplot() +
  geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0, dodge.width = 0.75)) +
  theme_classic() +
  facet_grid(~class, scales = "free", drop = TRUE) + 
  labs(fill = "Grazing regime", x = "Habitat", y = "Count") +
  scale_fill_manual(values = c("#77BFBC", "#EB956A"), labels = c("Winter", "Year-round")) +
  theme(panel.spacing = unit(.0, "lines"),
        panel.border = element_rect(color = "black", fill = NA, size = 1), 
        strip.background = element_rect(color = "black", size = 1),
        axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("../../3_output/2_graphs/seedling_counts.svg", height = 4, width = 5)
```

The models (GLMMs) below model numbers of individuals against "Country" and "Habitat" (called "section" in the data set). To reduce the complexity of the model and since I am not interested in relationships between tree classes, I ran this model over subsets containing data for each tree class separately (i.e., seedling, sapling, adult tree, or dead tree). This data set includes 0 when no individuals of said classes were found in a given section of the transects. 

**Seedlings**
```{r}
# seedlings
d <- tree_transect_data_full %>% 
  filter(class == "seedling") 

#fit.a <- glm(n ~ country, family = poisson, data = d)
#summary(fit.a)
#anova(fit.a)
#AIC(fit.a)

#fit.b <- glmer(n ~ country + (1 | transect), family = poisson, data = d) 
#summary(fit.b)
#anova(fit.b)
#AIC(fit.b)
#r.squaredGLMM(fit.b)

#fit.c <- glm.nb(n ~ country, data = d)
#summary(fit.c)
#anova(fit.c)
#AIC(fit.c)

#fit.d <- glmer.nb(n ~ country + (1 | transect), data = d) 
#summary(fit.d)
#anova(fit.d)
#AIC(fit.d)
#r.squaredGLMM(fit.d)

#fit <- glmer(n ~ country * section + (1 | transect), family = poisson, data = d) 
#summary(fit)
#anova(fit)
#AIC(fit)
#r.squaredGLMM(fit)

#fit1 <- glmer.nb(n ~ country * section + (1 | transect), data = d) 
#summary(fit1)
#anova(fit1)
#AIC(fit1)
#r.squaredGLMM(fit1)

fit2 <- glm(n ~ country * section, family = poisson, data = d)
summary(fit2)
anova(fit2)
AIC(fit2)
dispersiontest(fit2)

#fit3 <- glm.nb(n ~ country * section, data = d)
#summary(fit3)
#anova(fit3)
#AIC(fit3)
```

```{r}
lsmeans(fit2, pairwise ~ country * section, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts
```
•	Higher numbers of seedlings were found in the forest (p = 0.0002) and the treeline (p = < 0.0001), but not in the tundra (p = 0.9405) in Norway in comparison to the forest in Finland . Slightly higher numbers of seedlings were found in the Norwegian treeline compared to the Finnish treeline, although this finding lacks statistical support (p = 0.0561).

**Saplings**
```{r}
# saplings
d <- tree_transect_data_full %>% 
  filter(class == "sapling") 

fit <- glmer(n ~ country * section + (1 | transect), family = poisson, data = d) 
summary(fit)
anova(fit)
AIC(fit)
r.squaredGLMM(fit)
```

```{r}
lsmeans(fit, pairwise ~ country * section, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts
```
•	Sapling numbers were consistently higher in Norway in all habitats (forest, p < 0.0001; treeline, p = 0.0002; tundra, p = 0.0137) when compared to Finland, suggesting that intense grazing regimes limit recruitment in Finland. 

**Adult trees**
```{r}
# adult trees
d <- tree_transect_data_full %>%  
  filter(class == "tree") 

fit <- glmer(n ~ country * section + (1 | transect), family = poisson, data = d) 
summary(fit)
anova(fit)
AIC(fit)
r.squaredGLMM(fit)
```

```{r}
lsmeans(fit, pairwise ~ country * section, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts
```

**Dead trees**
```{r}
# dead
d <- tree_transect_data_full %>% 
  filter(class == "dead") 

fit <- glmer(n ~ country * section + (1 | transect), family = poisson, data = d) 
summary(fit)
anova(fit)
AIC(fit)
r.squaredGLMM(fit)
```

```{r}
lsmeans(fit, pairwise ~ country * section, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts
```

**Relationship between trees and dung counts**

```{r echo = FALSE}
facet_order <- c("seedling", "sapling", "tree", "dead")

tree_transect_data_full %>% 
  rename(habitat = section) %>% 
  inner_join(transect_data) %>%  
  mutate(class = factor(class, levels = facet_order)) %>% 
  ggplot(aes(x = total_dung, y = n, color = country)) +
  geom_point() +
  facet_wrap(~ class)
```

```{r}
fit <- tree_transect_data_full %>% 
  rename(habitat = section) %>% 
  inner_join(transect_data) %>%  
  filter(class == "seedling") %>% 
  glmer(n ~ total_dung + (1 | transect), family = poisson, data = .) 

summary(fit)
anova(fit)
```
•	This is corroborated by the significant negative relationship between dung counts and numbers of seedlings (GLMM, F1 = 6.03, p = 0.0146) and saplings (GLMM, F1 = 16.34, p = 0.0009) found in the transects. 

```{r}
fit <- tree_transect_data_full %>% 
  rename(habitat = section) %>% 
  inner_join(transect_data) %>%  
  filter(class == "sapling") %>% 
  glmer(n ~ total_dung + (1 | transect), family = poisson, data = .) 

summary(fit)
anova(fit)
```

```{r include = FALSE}
fit <- tree_transect_data_full %>% 
  rename(habitat = section) %>% 
  inner_join(transect_data) %>%  
  filter(class == "sapling") %>% 
  glmer(n ~ slope  (1  | transect), family = poisson, data = .) 

summary(fit)
anova(fit)
```

```{r}
tree_transect_data_full %>% 
  rename(habitat = section) %>% 
  inner_join(transect_data) %>%  
  ggplot(aes(x = aspect, y = total_dung, color = country)) +
  geom_point()

fit <- transect_data %>% 
  lmer(aspect ~ country + (country | transect) + (1  | habitat), data = .) 
summary(fit)
anova(fit)
```

## Adult tree morphology
**numbers of stems**
```{r}
tree_data$country <- factor(tree_data$country, levels = c("Storfjord", "Malla"))

tree_data %>%
  na.omit() %>%
  ggplot(aes(class, n_stems, fill = country)) + 
  geom_boxplot() +
  #geom_jitter(aes(), alpha = 0.3, position = position_jitterdodge(jitter.width = 0.1, dodge.width = 0.75)) +
  scale_fill_manual(values = c("#77BFBC", "#EB956A"), labels = c("Winter", "Year-round")) +
  labs(y = "number of root shoots", fill = "Grazing regime") +
  theme_classic() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1))

ggsave("../../3_output/2_graphs/root_shoots.svg", height = 4, width = 4)
```

```{r}
fit <- tree_transect_data %>% 
  rename(habitat = section) %>% 
  full_join(transect_data) %>%  
  filter(class != "dead") %>% 
  lmer(mean_n_stems ~ country * class + (country | transect) + (1 | habitat), data = .) 

summary(fit)
anova(fit)
lsmeans(fit, pairwise ~ country * class, lmer.df = "Satterthwaite", adjust = "tukey")$contrasts
```

**height and diameter**
```{r}
tree_data %>%
  filter(class == 'tree') %>%
  ggplot(aes(country, height, color = country)) + 
  geom_boxplot() 
```

**height to diameter ratio**
adult trees
```{r}
tree_data %>%
  filter(class == 'tree') %>%
  ggplot(aes(height, diameter, color = country)) + 
  geom_point() 
```

adult trees (per transect)
```{r}
tree_data %>%
  filter(class == 'tree') %>%
  ggplot(aes(height, diameter, color = country)) + 
  geom_point() +
  geom_smooth(method = 'lm') +
  facet_wrap(~transect)
```

saplings
```{r}
tree_data %>%
  filter(class == 'sapling') %>%
  ggplot(aes(height, diameter, color = country)) + 
  geom_point() +
  geom_smooth(method = 'lm')
```

saplings (per transect)
```{r}
tree_data %>%
  filter(class == 'sapling') %>%
  ggplot(aes(height, diameter, color = country)) + 
  geom_point() +
  #geom_smooth(method = 'lm') +
  facet_wrap(~transect)
```

seedlings
```{r}
tree_data %>%
  filter(class == 'seedling') %>%
  ggplot(aes(height, diameter, color = country)) + 
  geom_point() #+
  #geom_smooth(method = 'lm') +
  #facet_wrap(~transect)
```


**growth form vs stem diameter**
```{r}
tree_data %>%
  filter(class == 'tree') %>%
  ggplot(aes(n_stems, diameter, color = country)) + 
  geom_point() +
  geom_smooth(method = 'lm') 
```

**trees along transect**
```{r}
tree_data %>%
  ggplot(aes(distance, height, color = country)) + 
  geom_point() +
  geom_vline(aes(xintercept = 0), linetype = 'dashed') +
  facet_wrap(~transect)
```

transect N8 in particular shows a more smooth transition from continuous forest to open tundra
```{r}
tree_data %>%
  filter(transect == 'N8') %>%
  ggplot(aes(distance, height)) + 
  geom_point() +
  geom_vline(aes(xintercept = 0), linetype = 'dashed')
```

**numbers of stems**
```{r}
tree_data %>%
  filter(class == 'tree') %>%
  ggplot(aes(n_stems, color = country, fill = country)) + 
  geom_histogram(alpha = 0.5, 
                position="identity") +
  facet_wrap(~country)
```
